
## MMA

![photo:[pixabay](https://pixabay.com/photos/mma-mixed-martial-arts-grachan-cage-2944759/)](images/mma.jpg)


```{r, echo=F, include=F}
# setting up data exported from jamovi
# NOTE the csv file provided on plos one is corrupted. You must open the jamovi file and export from there to rds

mma.wide <- read_rds('data/mma/mma.rds') %>%
  mutate(person = row_number()) %>%
  rename(FI = `Wingate_fatigue index [%]`) %>%
  mutate(FI.c = FI - mean(FI))

# see stats section in paper
prerowstodrop <- mma.wide %>% select(starts_with('RT_before_')) %>% complete.cases()
postrowstodrop <- mma.wide %>% select(starts_with('RT_after_')) %>% complete.cases()
sum(!prerowstodrop)
sum(!postrowstodrop)

mma.repeated <- mma.wide %>%
  select(person, matches('RT_|MT_')) %>%
  pivot_longer(-person, names_to=c("outcome", "condition", "trial"), names_sep="_")

mma <- mma.repeated %>%
  pivot_wider(id_cols=c("person", "condition", "trial"), names_from=outcome, values_from=value) %>%
  left_join(., mma.wide %>% select(person, FI, FI.c)) %>%
  mutate(stimulus = as.numeric(trial)) %>%
  mutate(condition=fct_rev(condition)) %>%
  # filter(trial>4) %>%
  # centering, see syntax file here: https://journals.plos.org/plosone/article/file?id=10.1371/journal.pone.0227675.s005&type=supplementary
  mutate(stimulus.c = stimulus - mean(stimulus))

#xtabs(~condition+condition_12, data=mma)

rowstokeep <- mma %>% select(trial,RT,MT,FI) %>% complete.cases()

mma.raw <- mma %>%
  mutate(included=rowstokeep) %>%
  mutate(included = ifelse(stimulus < 5, FALSE, included)) %>%
  filter(included==TRUE)

mma.raw %>% select(person, condition, RT, stimulus) %>%
  write_csv('data/mma_bw.csv')

mma  <- read_csv('data/mma_bw.csv')
```


The following data are a tidied-up version of the data made available by the authors of the MMA paper we discussed in class [@pavelka2020acute]. The `included` column tells whether the row was included in the authors' analysis from the paper.

```{r}

```



:::{.exercise}

- Import the tidied the data (from [this file](data/mma_bw.csv))

- Fit the model described in the paper (in their supplement the authors reveal that `stimulus` and `FI` were mean-centered).

`r hide("How do I mean-centre things?")`

```{r}
mma.c <- mma %>%
  mutate(FI.c = FI - mean(FI), stimulus.c = stimulus - mean(stimulus))
mma.c %>% head
```

`r unhide()`


`r hide("Show the model")`

```{r}
(mma.m1 <- lmer(RT ~ 1 + condition + stimulus.c + FI.c + condition:stimulus.c + (1|person), data=mma.c))
```

`r unhide()`


- Look at the effect of `condition` in your model. Does this match the authors' report that "POST RTs were found significantly higher than PRE (mean difference = 4.031 ms, SE = 1.726"?

`r hide("Show answer")`

Yes: the `conditionbefore` coefficient from your model should have been -4.03 also. The direction is reversed because the authors' encoded their model slightly differently (their parameter would have been called `conditionafter`) but the meaning is the same.

To see the SE, use `summary` and `coef` together:

```{r}
summary(mma.m1) %>% coef() %>% pander()
```


This is 0.17660, which is what the authors report.


`r unhide()`



- Are the reported findings robust to minor changes in the model fitting? For example, if parameters are removed or if the excluded data is incorporated?

`r hide("Show details")`


This model, excluding fatigue (`FI`), predicts the same 4ms difference:

```{r}
mma.m2 <- lmer(RT ~ 1 + condition * stimulus.c + (stimulus.c|person), data=mma.c)
summary(mma.m2) %>% coef() %>% pander()
```

However this model fit includes all the excluded data and predicts a 2ms difference in the other direction, and is not statistically significant:

```{r}
mma.all  <- read_csv('data/mma_bw.csv') %>%
  mutate(stimulus.c = stimulus-mean(stimulus), FI.c = FI-mean(FI), )
mma.m3 <- lmer(RT ~ 1 + condition + stimulus.c + FI.c + condition:stimulus.c +  (1 | person), data=mma.all)
mma.m3 %>% summary() %>% coef() %>% pander()
```


`r unhide()`


- How should we interpret these findings, based on the replications you have made?


:::





















## Extensions

### Factorial anova in mixed models

The file [journal.pone.0226387.s001.sav](data/journal.pone.0226387.s001.sav) contains data from @kemps2019cognitive.
These have been reshaped to long format and tidied up in [data/drinks.csv](data/drinks.csv).

The study explored the effect of an cognitive bias modification intervention, and one of the outcomes was Attentional Bias in a dot-probe task.

There were two experimental conditions (between subjects) and participants were tested twice. The original paper reported results from a mixed model in which `Time` was a repeated/within-subjects factor.

```{r echo=F, include=F}
drinks_ <- haven::read_sav('data/journal.pone.0226387.s001.sav') %>%
  mutate_if(~class(.)=="haven_labelled", funs(as_factor))
# drinks_ %>% glimpse

longtmp <- drinks_ %>%
  select(ParticipantID, Att_bias_pre_training, Att_bias_post_training) %>%
  pivot_longer(-ParticipantID, names_pattern = "Att_bias_(pre|post)_training")
class(drinks_$Gender)
drinks <- left_join(longtmp, drinks_, by="ParticipantID") %>%
  select(Participant=ParticipantID, Condition, Age, Gender, Time=name, AttBias=value) %>%
  mutate(Time=Hmisc::capitalize(Time))
drinks %>% write_csv('data/drinks.csv')
drinks  <-  read_csv('data/drinks.csv')
```

```{r}
drinks %>% head %>% pander("6 rows from the `drinks` dataset")
```


Plotting the data indicates a Time*Condition interaction:

```{r, echo=F}
drinks %>%
  ggplot(aes(fct_rev(Time), AttBias, color=factor(Condition),
             group=factor(Condition))) +
  stat_summary(fun.data=mean_se, position=position_dodge(width=.1)) +
  stat_summary(geom="line", fun.data=mean_se, position=position_dodge(width=.1)) +
  xlab("Time") + ylab("Attentional Bias (ms change)") + scale_color_discrete("Condition")

ggsave("images/drinksplot.png", width=4, height=3)
```



:::{.exercise}

1. Run a random intercepts model which would test the relationships shown in the plot above

`r hide("Solution")`

```{r}
(drinks.m1 <- lmer(AttBias ~ Condition * Time + (1|Participant), data=drinks))
```

`r unhide()`


2. Use the `anova` command to test the interaction of Condition and Time


`r hide("Solution")`

```{r}
drinks.m1 %>% anova
```

There does appear to be an interaction.

`r unhide()`


:::














# Follow-up tests

To run follow-up tests comparing individual cells in the design we can use another R package called `emmeans`.


```{r}
library(emmeans)
```


The `emmeans` function (which is inside the `emmeans` package we just loaded), can calculate means and confidence intervals for each cell in the design:

```{r}
emmeans(drinks.m1, ~Condition*Time)
```

In the code above we used the `emmeans` function, and gave it the saved random intercept model as input. In the second input (where is says `~Condition*Time`) we are using a formula to describe which part of the design we would like means for.

Helpfully, you can also plot the results of this function, so the folliwing is a shortcut for writing a ggplot command yourself:

```{r}
plot(emmeans(drinks.m1, ~Condition*Time)) + xlab("Predicted value")
```


We can also save the results of `emmeans` and send it to the `contrast` function. This produces t tests for each pairwise comparison:

```{r}
savedmeans1 <- emmeans(drinks.m1, ~Condition*Time)
contrast(savedmeans1, method='pairwise', adjust='none')
```

And we can add `adjust="bonferonni"` or `adjust="tukey"` method to adjust for multiple comparisons

```{r}
contrast(savedmeans1, method='pairwise', adjust='tukey')
```


If we want to compare pairs of groups, we can adjust the formula we pass to the `emmeans` function. In the example below I compare Conditions averaged across both time periods:

```{r echo=F, include=F}
emmeans(drinks.m1, ~Condition)
```


```{r}
savedmeans2 <- emmeans(drinks.m1, ~Condition)
contrast(savedmeans2, method='pairwise')
```

In this example we compare the Pre and Post measurement times *within* each group (here the bar symbol, `|`, is read as 'within'):

```{r}
savedmeans3 <- emmeans(drinks.m1, ~Time|Condition)
contrast(savedmeans3, method='pairwise')
```


:::{.tip}

##### Useful `emmeans` examples


```{r, echo=T, eval=F}
# get the means of all combinations of factors A and B
emmeans(model, ~A+B)

# menas the same as the line above
emmeans(model, ~A*B)

# show all pairwise contrasts for all combinations of A and B
contrast(emmeans(model, pairwise~A+B), pairwise)

# show tests of the effect of A within each level of B
contrast(emmeans(model, pairwise~A|B), pairwise)

# show the confidence interval for the tests above:
confint(contrast(emmeans(model, pairwise~A|B), pairwise))

```

:::




:::{.exercise}

- Adjust the code using `emmeans` and `contrast` above to compare Pre vs Post scores (averaged across Condition)
- Compare the conditions *within* each time period.

`r hide("Solution")`

```{r}
# compare times averaged across conditions
contrast(emmeans(drinks.m1, ~Time), method='pairwise')
# compare conditions within each time
contrast(emmeans(drinks.m1, ~Condition|Time), method='pairwise')
```


`r unhide()`
:::

