---
output: pdf_document
title: Schools activity
---

```{r setup, include=FALSE}
library(tidyverse)
library(pander)
library(broom)
library(webex)

options(scipen=4, digits=4)
knitr::opts_chunk$set(
  echo = T,
  include = T, 
  cache=T,
  comment="",
  warning=F,
  message = F
)

theme_set(theme_gray(base_size = 16))
```





Imagine you work for the UK Department for Education (DfE). 

**Your task is to find out whether a new educational intervention is effective**. The new intervention has been rolled out in 1000 randomly-selected schools in the UK.

- You have a budget of around £10,000. 
- The cost to visit a single school is £100
- The cost to measure each pupil is £10
- The average size of a school in the UK is roughly [250](https://www.theguardian.com/education/2013/may/17/primary-schools-supersize).

Would you prefer to collect data from:

```{r echo=F, results='asis'}
budget <- 10000
schoolcost  <- 100
kidcost <- 10
p  <- seq(.9,.1, -.1) # prop spent on schools
schools = as.integer(p*budget / schoolcost)
kids = as.integer((1-p)*budget / kidcost)
expand.grid(schools=seq(3,200,10), kids=1:1000) %>% 
  as.tibble() %>%
  mutate(
    cost = kids*kidcost+schools*schoolcost)%>%
  filter(cost == budget & kids > schools)%>%
  group_by(schools)%>%
  arrange(-cost)%>%
  slice(1:1)%>%
  mutate(
    s = sprintf("%s children across %s schools", 
      kids,
    schools
    )) %>%
pull(s) %>% pander::pandoc.list()
```

These options all cost around £10,000.

Discuss the problem as a small group. Identify the different tradeoffs involved.



